---
title: "lab_02"
author: "derek willis"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: https://worldpopulationreview.com/us-counties/maryland.

**17,099 voters appear to be new registrants, with Montgomery County seeing the most seemingly new voters.**

```{r}

selected_md_voters_24 <- md_voters_2024 |> select(FIRSTNAME, LASTNAME, STATE_REGISTRATION_DATE, COUNTY_REGISTRATION_DATE, COUNTY)


selected_md_voters_24 |>
  filter(STATE_REGISTRATION_DATE >= "2024-07-21" & COUNTY_REGISTRATION_DATE >= "2024-07-21" & STATE_REGISTRATION_DATE == COUNTY_REGISTRATION_DATE) |>
  group_by(COUNTY)

```

```{r}

selected_md_voters_24 <- md_voters_2024 |> select(FIRSTNAME, LASTNAME, STATE_REGISTRATION_DATE, COUNTY_REGISTRATION_DATE, COUNTY)


selected_md_voters_24 |>
  filter(STATE_REGISTRATION_DATE >= "2024-07-21" & COUNTY_REGISTRATION_DATE >= "2024-07-21" & STATE_REGISTRATION_DATE == COUNTY_REGISTRATION_DATE) |>
  group_by(COUNTY) |>
  count(COUNTY) |>
  arrange(desc(n))

```


**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**There was a large increase in county voter registrations in the five days immediately following July 21st, but this pattern was pretty standard throughout the dataset. The dates that experience this 5 day increase in registrations are all weekdays. The increase might be attributed to people most often registering to vote on weekdays. On average, however, the overall number of voter registrations were higher in the 2 weeks following Biden's announcement than they were prior.**

```{r}

md_voters_2024 <- md_voters_2024 |> mutate(week(COUNTY_REGISTRATION_DATE))

```


```{r}

md_voters_2024 |>
  group_by(COUNTY_REGISTRATION_DATE, `week(COUNTY_REGISTRATION_DATE)`) |>
  summarise(
    voters = n()
    ) |>
  arrange(desc(COUNTY_REGISTRATION_DATE))

```


**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?
**July had the highest number of voter registrations. This helps shed light on Q2 by showing how registration rates vary by month.** 

```{r}

md_voters_2024 |>
  mutate(month = floor_date(COUNTY_REGISTRATION_DATE, "month")) |>
  group_by(month) |>
  summarise(voters = n()) |>
  arrange(desc(voters))

```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**Central Maryland saw the largest number of seemingly new voters with 7583 people who applied on or after July 21, 2024. The party with the largest number of new voter applicants was the democratic party with 7650 registrations on or after July 21, 2024.** 

```{r}


stand_md_voters_2024 <- md_voters_2024 |>
  mutate(
    COUNTY = str_remove_all(COUNTY, "'")
)


```


```{r}

stand_md_voters_2024 |>
  mutate(
    REGION = case_when(
      COUNTY == "Allegany" | COUNTY == "Washington" | COUNTY == "Garrett" ~ "Western Maryland",
      COUNTY == "Frederick" | COUNTY == "Montgomery" | COUNTY == "Prince Georges" ~ "Capital Region",
      COUNTY == "Anne Arundel" | COUNTY == "Baltimore City" | COUNTY == "Baltimore County" | COUNTY == "Carroll" | COUNTY == "Harford" | COUNTY == 
        "Howard" ~ "Central Maryland",
      COUNTY == "Calvert" | COUNTY == "Charles" | COUNTY == "Saint Marys" ~ "Southern Maryland",
      COUNTY == "Kent" | COUNTY == "Queen Annes" | COUNTY == "Talbot" | COUNTY == "Caroline" | COUNTY == "Dorchester" | COUNTY == "Wicomico" | COUNTY == "Somerset" | COUNTY == "Worcester" | COUNTY == "Cecil" ~ "Eastern Shore"
    )
  ) |>
  filter(STATE_REGISTRATION_DATE >= "2024-07-21" & COUNTY_REGISTRATION_DATE >= "2024-07-21" & STATE_REGISTRATION_DATE == COUNTY_REGISTRATION_DATE) |>
  group_by(REGION) |>
  summarise(voters = n()) |>
  arrange(desc(voters))

```
```{r}

stand_md_voters_2024 |>
  filter(STATE_REGISTRATION_DATE >= "2024-07-21" & COUNTY_REGISTRATION_DATE >= "2024-07-21" & STATE_REGISTRATION_DATE == COUNTY_REGISTRATION_DATE) |>
  group_by(PARTY) |>
  summarise(voters = n()) |>
  arrange(desc(voters))

```


**Q5** What's an additional question you could ask of this data?

**You could use this data to ask which months see higher rates of new voter registration, and then use those results to find why those months see more voter registrations than others. If you had access to previous years' data, you could also compare this to non-election years or other election years.** 
