---
title: "lab_02"
author: "derek willis"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)

```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: https://worldpopulationreview.com/us-counties/maryland.

**17206**

```{r}
library(dplyr)

md_voters_2024 <- md_voters_2024 %>%
  mutate(COUNTY_REGISTRATION_DATE = as.Date(COUNTY_REGISTRATION_DATE, format = "%Y-%m-%d"))

new_voters <- md_voters_2024 %>%
  filter(COUNTY_REGISTRATION_DATE > as.Date("2024-07-21"))

nrow(new_voters)
```

**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**From July 21-Aug 7, voters groups decreased from around 7,000 to around 3,000. (I used an AI assistant to help with some of the code here)**

```{r}
md_voters_2024 <- md_voters_2024 %>%
  mutate(week_number = week(COUNTY_REGISTRATION_DATE))
voters_by_week <- md_voters_2024 %>%
  group_by(week_number) %>%
  summarise(voter_count = n())
weeks_of_interest <- voters_by_week %>%
  filter(week_number %in% c(30, 31))
# Get the overall distribution of registrations by week
voter_distribution <- ggplot(voters_by_week, aes(x = week_number, y = voter_count)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  theme_minimal() +
  labs(title = "Voter Registrations by Week", x = "Week Number", y = "Number of Voters")

print(voter_distribution)

```

**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?
**July has the most voter registrations. This does shed more light becasue there cannot be more after that point, so the decrease makes sense!** 

```{r}

df <- read.csv("lab_02.rds", stringsAsFactors = FALSE)
df$STATE_REGISTRATION_DATE <- as.Date(df$STATE_REGISTRATION_DATE, format="%Y-%m-%d")
df$Registration_Month <- month(df$STATE_REGISTRATION_DATE, label = TRUE)
most_common_month <- names(which.max(table(df$Registration_Month)))
```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**Capital Region Democrats account for the most new voters.** 

```{r}
library(dplyr)

df <- read.csv("lab_02.rds", stringsAsFactors = FALSE)

df$STATE_REGISTRATION_DATE <- as.Date(df$STATE_REGISTRATION_DATE, format="%Y-%m-%d")

df <- df %>%
  mutate(region = case_when(
    COUNTY %in% c("Allegany", "Garrett", "Washington") ~ "Western Maryland",
    COUNTY %in% c("Montgomery", "Frederick", "Prince George's") ~ "Capital Region",
    COUNTY %in% c("Baltimore", "Baltimore City", "Carroll", "Harford", "Howard", "Anne Arundel") ~ "Central Maryland",
    COUNTY %in% c("Calvert", "Charles", "St. Mary's") ~ "Southern Maryland",
    COUNTY %in% c("Caroline", "Cecil", "Dorchester", "Kent", "Queen Anne's", "Somerset", "Talbot", "Wicomico", "Worcester") ~ "Eastern Shore",
    TRUE ~ "Unknown"
  ))

df$Registration_Month <- month(df$STATE_REGISTRATION_DATE, label = TRUE)

region_counts <- df %>%
  group_by(region) %>%
  summarise(new_voters = n()) %>%
  arrange(desc(new_voters))

print("New Voters by Region:")
print(region_counts)

party_counts <- df %>%
  group_by(PARTY) %>%
  summarise(new_voters = n()) %>%
  arrange(desc(new_voters))

print("New Voters by Party:")
print(party_counts)

region_party_counts <- df %>%
  group_by(region, PARTY) %>%
  summarise(new_voters = n()) %>%
  arrange(region, desc(new_voters))

print("Most Common Party in Each Region:")
print(region_party_counts)

```

**Q5** What's an additional question you could ask of this data?

**You could ask which age group is registering to vote the most commonly.** 
