---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2025_spring")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```

```{r}

# print will take a string and return it in the terminal.
print("Hello World.")

```

```{r}

# This is a comment inside of a codeblock. Comments doesn't run when we run the codeblock.
# Adding this command is a good way to start your class reference notebook.

```

## read/load data
```{r}

NAME(no hyphens) <- read_rds("data/NAME.rds")

```

```{r}

NAME(no hyphens) <- read_csv("data/NAME.csv")

```

#preland 2
```{r}

#lubridate: a system that makes it to R Studio can read, extract, and work with dates and time data better.

# head(DATASETNAME): simplifies the presentation of the dataset.

# summary(DATASETNAME): provides a table summarizing the overall dataset.

# colnames(DATASETNAME):displays the column names.

#glimpse(DATASETNAME): provides an overall glimpse of the data (columns, rows and what data exists in those categories.)


```


#better ordering
```{r}

#creates a new column/variable with `mutate` by using preexisting data, showing the percentage of votes made per candidate on election day. Then arranges the data by the percent_election_day variable.

primary_24 |>
  select(office_name, office_district, candidate_name, party, county_name, election_day, votes) |>
  mutate(
  percent_election_day = (election_day/votes)*100
)  |> 
  arrange(desc(percent_election_day))

```

```

