---
title: "Reverse Engineering Project"
author: "Yugo Hines"
date: "March 12 2025"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## Introduction 

In this notebook, we are reverse engineering the story, [STORY NAME](STORY URL)

## Load libraries

Loading required libraries for this analysis.

```{r}

library(tidyverse)
library(lubridate)
library(refinr)
library(janitor)
library(dplyr)

```


## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data
donors <- read_csv("data/donors0103-0701.csv")

glimpse(donors)

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 


```
##data cleaning vvvvv

```{r}
donors <- donors |>
  mutate(
    amount = str_remove_all(amount, "\\$"),
    amount = str_remove_all(amount, ",")
  )

glimpse(donors)

```


```{r}

donors <- donors |>
  mutate(amount = as.numeric(amount)) |>
  mutate(date = mdy(date)) |>
  mutate(committee_zip = str_sub(committee_zip, start=1L, end=5L)) 
  
glimpse(donors)

```


## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: Add in the week before session began, and the two businessmen gave more than $87,000 to Republican legislators and the North Carolina Republican Party, a key cog in the campaign machine this year as the two major parties fight for control of the General Assembly in November's elections.

* **Analysis summary**:  Before we could confirm the findings, we had to mutate several columns; zip codes needed to be five digits, `date` needed to have the <date> datatype, `amount` needed to be the <dbl> datatype, and `name-clean` needed to be renamed to `name_clean`. After cleaning the data, we were able to confirm the findings by filtering amounts donated by the two donors named in the article. The results confirmed that the George Strobel and Robin Delmer did donate over $87,000.


```{r}

donors <- donors |>
  rename(
    name_clean = "name-clean"
  )

```

```{r}

# Put code to reverse engineer sentence here
donors |>
  filter(name_clean == "GEORGE LEWIS STROBEL II" | name_clean == "ROBIN E DELMER") |>
  group_by(name_clean) |>
  summarise(amount = sum(amount)) |>
  arrange(desc(amount))

# Display results of code below this codeblock


```


### Sentence 2

* **Sentence text**: This year's legislative session also saw significant donations from hog farmers and beer distributors, both industries with important legislation before the General Assembly. Senate President Pro Tem Phil Berger's campaign alone raised nearly $222,000 during session. Twenty-three people gave him the maximum allowed: $5,200.

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 3

* **Sentence text**: All told during the six-week General Assembly session, more than $1.1 million flowed into state legislators' campaign accounts. That's a conservative figure, reached by backing out money that simply flowed from one politician's campaign to another and money elected officials put into their own campaigns.

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 4

* **Sentence text**: Add in donations to the two main political parties and to a handful of other large fundraising vehicles for state legislative races, and the in-session giving topped $3.4 million. Gov. Roy Cooper's main campaign account brought in another $407,000 during session.

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

* **Sentence text**: “A WRAL News analysis of campaign finance filings also turned up $16,000 in donations to legislative campaigns during session from 15 PACs with a lobbyist. Each would appear to be a violation of state fundraising rules, but spokespeople for most of the PACs, or the companies they represent, said there was a simple explanation: The checks were written before session.”

* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-