---
title: "lab_03"
author: "Derek Willis"
date: "2024-08-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## Load data

```{r}
# Load MD slice of FEC House candidate contribution data
md_contribs <- read_rds("data/lab_03.rds")
```

## Fix header issues

You'll need to fix one issue with the column names using janitor. Write a code block to do that, making sure that you save the results to the same dataframe.

```{r}

library(janitor)

md_contribs_clean <- md_contribs |>
  clean_names() |>
    mutate(contributor_zip = str_sub(contributor_zip, start=1L, end=5L)) |>
  mutate(contribution_receipt_date = mdy(contribution_receipt_date))

md_contribs_clean

```


## Answer questions

Q1: Which seven cities in Maryland had the largest *number* of contributions from donors? How many did each have? To answer this question, you'll have to clean the messy "contributor_city" field.

Note: To answer this question, you'll need to see every variation of the "contributor_city" for the top seven cities, which you should identify through some initial analysis. Then find variations of them and change those variations into the most common version.

**The top seven cities for contribution amounts are Bethesda, Chevy Chase, Potomac, Baltimore, Annapolis, Silver Spring and Rockville.**

```{r}

two_md_contribs_clean <- md_contribs_clean |>
  mutate(contributor_city = case_when(
    contributor_city == "4615 NORTH PARK AVE APT. 1410" ~ "CHEVY CHASE",
    contributor_city == "AAAAAA OXON" ~ "OXON HILL",
    contributor_city == "ANAPOLIS" ~ "ANNAPOLIS",
    contributor_city == "BALRIMORE" ~ "BALTIMORE",
    contributor_city == "BLATIMORE" ~ "BALTIMORE",
    contributor_city == "BELAIR" ~ "BEL AIR",
    contributor_city == "COLOMBIA" ~ "COLUMBIA",
    contributor_city == "DAVISDSONVILLE" ~ "DAVIDSONVILLE",
    contributor_city == "ELLICOTT CITY" ~ "ELLICOTT",
    contributor_city == "GAITHERBURG" ~ "GAITHERSBURG",
    contributor_city == "LUTHERVILLE" ~ "LUTHERVILLE-TIMONIUM",
    contributor_city == "LUTHERVILLE TIMO" ~ "LUTHERVILLE-TIMONIUM",
    contributor_city == "LUTHERVILLE TIMONI" ~ "LUTHERVILLE-TIMONIUM",
    contributor_city == "LUTHERVILLE-TIMONIUM" ~ "LUTHERVILLE-TIMONIUM",
    contributor_city == "LUTHVLE TIMON" ~ "LUTHERVILLE-TIMONIUM",
    contributor_city == "TIMONIUM" ~ "LUTHERVILLE-TIMONIUM",
    contributor_city == "MARRIOTSVILLE" ~ "MARRIOTTSVILLE",
    contributor_city == "MT.AIRY" ~ "MT. AIRY",
    contributor_city == "NEW MARKED" ~ "NEW MARKET",
    contributor_city == "N BETHESDA" ~ "NORTH BETHESDA",
    contributor_city == "OWINS MILLS" ~ "OWINGS MILLS",
    contributor_city == "POTAMIC" ~ "POTOMAC",
    contributor_city == "RIVERDALE" ~ "RIVERDALE PARK",
    contributor_city == "SILVER SPRINGS" ~ "SILVER SPRING",
    contributor_city == "ST MARYS CITY" ~ "SAINT MARYS CITY",
    contributor_city == "ST MICHAELS" ~ "SAINT MICHAELS",
    contributor_city == "ST. CHARLES" ~ "SANT CHARLES",
    contributor_city == "UPPR MARLBORO" ~ "UPPER MARLBORO",
    TRUE ~ contributor_city
  )
  )

```


```{r}

two_md_contribs_clean |>
  group_by(contributor_city) |>
  summarise(contribs = sum(contribution_receipt_amount)) |>
  arrange(desc(contribs))

```

Q2: Find the 5-digit ZIP code in Maryland that had the highest total amount of contributions (by dollar value). You will need to use mutate to ensure that you have a zip code column with the first five characters from the existing zip code column.

Where is that zip code and what kind of area is it, demographically? Which committee is the largest recipient of money from that zip code? How many of the top five recipients are Maryland candidates?

*The five digit zip code is 20817. The area has a large demographic of white adults. *

```{r}
#mutated contributor_zip into a 5 digit zip code on line 42
two_md_contribs_clean |>
  group_by(contributor_zip) |>
  summarise(zip_contribs = sum(contribution_receipt_amount)) |>
  arrange(desc(zip_contribs))

```

```{r}

two_md_contribs_clean |>
  filter(contributor_zip == 20817) |>
  group_by(committee_name) |>
  summarise(com_contribs = sum(contribution_receipt_amount)) |>
  arrange(desc(com_contribs))

```


Q3: Let's look at top individual donors. Write code to generate a list of top donors using name and city, adding up the total amount of contributions. What does the original data say about the top two donors - who are they and what do they do? Do you see anything in the list of top 10 donors that would make you uncertain about the accuracy of these results?

*The data says that the top two donors are "Elfreth for Maryland" in Annapolis and "Johnny O for Congress" in Baltimore. Both committees are political committees that are backing two specific politicians for congress. One thing I'm unsure of is the fact that some committees have multiple branches in different city. For example, "Jamie Raskin for Congress" has a branch in both Chevy Chase and Bethesda. Would it be better to group these two branches together, or is it better to separate them by city? I thinkthe type of story the data is being used for would determine whether or not branches should be grouped together. *

```{r}

two_md_contribs_clean |>
  group_by(committee_name, contributor_city) |>
  summarise(commit_contribs = sum(contribution_receipt_amount)) |>
  arrange(desc(commit_contribs))

```
