
---
title: "pre_lab_01.Rmd"
author: "Derek Willis & Ryan Little"
date: "2025-02-01"
output: html_document
---

```{r setup, include=FALSE}
# this chunk is for setup and global options
knitr::opts_chunk$set(echo = TRUE)

#turn off scientific notation
options(scipen=999)
```

## About this notebook

This notebook contains code and explanatory text that your should review and run as you read through chapters 13 and 10 of the course textbook, "Data Journalism with R and the Tidyverse". Answer questions and edit the document as directed.

Running this notebook will help you understand key data analysis methods and concepts that you will put into practice during this week's lab. When you come across a new command or verb in R, you should add it to your class reference notebook file along with what it's used for.

When you are finished running the code in this notebook, you will push changes to your course GitHub repo, and upload the link to ELMS as instructed.

You also should update your class reference notebook with code examples and what they mean. When you come across a new concept or R verb, put it in your reference notebook along with a sentence about what it does. For example:

```{r}
library(tidyverse)
```

The library function loads an R library such as the tidyverse, providing access to the verbs and functions that library contains. You have to load a library before you can use its functions.

## Chapter 13 [https://thescoop.org/data_journalism_book/r-basics.html]

### Task 1: The console

**Task** In the console (look below), type 2+2 and hit enter. What does it return? Type your answer below. **Answer:**

### Task 2: Variables

**Task** Put the following code -- which stores the numeral 2 as a variable called number -- in the console and hit enter.

```{r}
number <- 2
```

Then type this code to multiply the variable called number by itself.

```{r}
number * number
```

What does it return? Type your answer below:

NOTE: **Answer: Type your answer inside the asterisks like this.** 

**Answer:**

### Task 3: Load libraries

**Task** Run the following code in the gray-colored codeblock below -- not in the console -- to load the tidyverse library. To run the code, click the little green play button (left facing arrow) at the top right of the codeblock. In Rmarkdown data notebooks, we write code inside of codeblocks, and explanatory text in the white area outside of it.

```{r}
# This is a comment inside of a codeblock. Comments doesn't run when we run the codeblock.
# Adding this command is a good way to start your class reference notebook.
library(tidyverse)
```

## Chapter 15 [https://thescoop.org/data_journalism_book/aggregates.html]

### Task 5: Load data

**Task** Load the UMD courses data by running the following codeblock. The data is contained in an .rds file, which is a binary format for R data.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
```

Once it loads, look at the environment window at right in RStudio for umd_courses It's next to a small blue circle with a white arrow in it. To the right of that, look at how many "obs" the dataset has ("obs" is short for observation, which is another way of saying rows or records). Look at how many variables it has (another way of saying columns or fields). Write the number of observations and variables below.

**Answer**

### Task 6: Glimpse data

**Task** Run the following codeblock to use glimpse() to get a sense of the data, some of the values in each field/variable/column, and the data types of each field/variable/column.

```{r}
glimpse(umd_courses)
```

In the space below, write the data type for the "id" column and the "seats" column (it's in light gray, offset by \< \>). 

**Answer: "id" column is <chr> and "seats" is <dbl>.**

### Task 7: Examine data "head"

**Task** Run the following codeblock to use head() to examine the first six rows of data. Explore the data by clicking the black arrow on the right side of the output table, allowing you to examine additional columns.

```{r}

head(umd_courses)
```

In the space below, write the name of the business from the first row of the "name" column. 

**Answer: Directed Research.**

### Task 8: Explore data with interactive dataframe

**Task** look at the environment window at right in RStudio for umd_courses. It's next to a small blue circle with a white arrow in it. Click on the umd_courses -- not the blue circle. It will open the dataframe as an interactive table. Scroll down to see additional rows. Scroll left to see additional columns. Click on a column name to sort by that column.

### Task 9: Group and count

So what if we wanted to know how many classes were offered in each term?

To do that by hand, we'd have to take each of the 79,366 individual rows (or observations or records) and sort them into a pile. We'd put them in groups -- one for each term -- and then count them. To do this, we'll take our dataset and we'll introduce a new operator: `|>`. The best way to read that operator, in my opinion, is to interpret that as "and then do this." The keyboard shortcut for this is cmd-shift-m on the Mac and ctrl-shift-m on Windows. Don't like that character? R now has a new one that does the same thing: `|>`.

**Task** Run the following codeblock:

```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n()
  )
```

Describe what you think each line of this code is doing below, based on the result:

**Answer: umd_courses identifies what file we're pulling data from. group_by(term) groups all classes into the term they were taken in, and the attattched summarise(count_classes =n() ) adds a new column that lists the number of classes in each term.**

### Task 10: Adding arrange to sort the data

**Task** Run the following codeblock:

```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(count_classes))
```

Compared to the previous codeblock, what has changed and how did that affect the result? **Answer: This codeblock also uses the "arrange(desc(count_classes))," which arranges the terms in descending order by the number of classes they have.**

### Task 11: Grouping by multiple columns

**Task** Run the following codeblock.

```{r}
umd_courses |>
  group_by(term, department) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(term, department)
```

Compared to the previous codeblock, how did the output change and what in the code changed it?

**Answer: This codeblock adds a second input for the group command, making it "group_by(term, department)." This adds separate columns for both term and department.**

### Task 12: Other summary functions

**Task** Run the following codeblock:

```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats)
  ) |>
  arrange(desc(total_seats))
```

Compared to the previous codeblock, what has changed and how did that affect the result? 
**Answer: This codeblock removed department as a group and summarizes with both count_classes and total_seats. This shows each term, the number of classes for that term, and the total number of seats available for that term. The results are arranged by descending number of total seats.**

### Task 13: different kinds of summarizations

**Task** Run the following codeblock to group the umd_courses dataframe by term. Then do the following summarizations for each term: count the number of courses, add up the total of all seats, and calculate the mean and median contribution number of seats. Then sort so the term with the most total seats is listed first.

```{r}
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
    median_seats = median(seats)
  ) |>
  arrange(desc(total_seats))
```

Compared to the previous codeblock, what has changed and how did that affect the result? What do the top terms have in common?
**Answer: In addition to the functions of the previous codeblock, the new codeblock uses the summarise function to calculate the mean number of seats and median number of seats per term. The top terms have very similar mean and median seat numbers.**

### Task 14: more aggregations

**Task** Run the following codeblock to group the umd_courses dataframe by the department. Then do the following summarizations for each group: count the number of courses, add up the total of all seats, calculate the mean and median number of seats and then add lines to calculate the smallest and largest seat numbers using the `min()` and `max()` functions. Then sort so the department with the largest total number of seats is listed first.

```{r}
umd_courses |>
  group_by(department) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
    median_seats = median(seats),
    min(seats),
    max(seats)
  ) |>
  arrange(desc(total_seats))
```

Compared to the previous codeblock, what has changed and how did that affect the result? How is the data sorted?

**Answer: This block adds the min(seats) and max(seats) commands, which calculate the minimum and maximum number of seats available in all of the classes per department.**

### Task 12: Questions

No code here; I want you to provide three questions that this data suggests to you. What would you like to ask it?

**Answer**

1.  Why do some functions need to be ended with a "|>" and some don't, specifically in the summarize command?

2.  What do data types like <chr> and <dbl> mean? 

3.  With classes that have a 0 seat count, does that mean the class wasn't held that term? If so, should we still count those classes towards other things? For example, total number of classes per department.