---
title: "lab_10"
author: "Derek Willis"
date: "2024-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   the tidyverse, janitor and tigris libraries
-   you might choose to use the tidycensus library for Q3

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load libraries
library(tidyverse)
library(sf)
library(janitor)
library(tigris)


```

Let's explore the election results from last night and make some maps!

## Questions

**Q1.** Make a county-level map of Maryland's results for U.S. Senate using `md_senate_county_24.csv` in the data folder, calculating the difference between Angela Alsobrooks's percentage of the total votes and Larry Hogan's percentage of the total votes and displaying that percentage difference on the map. What is the story here, and what is the county with the most interesting results?

**The map shows the percentage difference between Alsobrooks and Hogan across Maryland’s counties. Areas with significant differences are clearly visible, with some counties strongly favoring one candidate over the other. The county with the most interesting results is PG county, which had the highest difference in votes between the two candidates.**

```{r}

# Reload senate data
senate_data <- read_csv("data/md_senate_county_24.csv") %>%
  clean_names()

# Calculate percentage for both candidates and the difference
senate_data <- senate_data %>%
  mutate(
    total_votes = alsobrooks + hogan,
    alsobrooks_pct = alsobrooks / total_votes * 100,
    hogan_pct = hogan / total_votes * 100,
    diff_percentage = alsobrooks_pct - hogan_pct
  )

# Load counties
us_counties <- counties(cb = TRUE)
md_counties <- us_counties %>%
  filter(STATEFP == "24")

# Join
senate_map_data <- md_counties %>%
  left_join(senate_data, by = c("NAME" = "county"))

ggplot(senate_map_data) +
  geom_sf(aes(fill = diff_percentage)) +
  scale_fill_gradient(low = "lavender", high = "purple") +
  labs(
    title = "Alsobrooks vs Hogan: percentage point difference by csounty",
    fill = "Percent Difference"
  )

```

**Q2.** Make a county-level map showing the difference between Donald Trump's county-level performance this year and Larry Hogan's, using percentages to compare the two as you did in Q1. You can use the dataframe you initially loaded for Q1, and you'll need to load the 2024 presidential results in the data folder and join them to it before proceeding. Are there any counties where Trump got a higher percentage than Hogan? How would you describe the map showing the Trump-Hogan difference?

Also answer this: is a map the best way to present this data? What else could you make that might convey more information?

**Trump was below Hogan in nearly every county, including some rural areas. A map shows geographic trends, sure, but a bar chart or even a line chart might better highlight the size of the differences. If you wanted to keep the focus on the counties you could do side by side tree maps of what percent of each county voted for hogan vs trump, keeping same counties in the same color across both charts.**

```{r}
pres_24 <- read_csv("data/md_pres_county_24.csv") %>%
  clean_names()

pres_24 <- pres_24 %>%
  mutate(
    total_votes = harris + trump + oliver + stein + kennedy,
    trump_pct = trump / total_votes * 100
  )

senate_data <- read_csv("data/md_senate_county_24.csv") %>%
  clean_names()

senate_data <- senate_data %>%
  mutate(
    total_votes = alsobrooks + hogan,
    hogan_pct = hogan / total_votes * 100
  )

senate_pres_data <- senate_data %>%
  left_join(pres_24 %>% select(county, trump_pct), by = "county") %>%
  mutate(trump_hogan_diff = trump_pct - hogan_pct)

us_counties <- counties(cb = TRUE)
md_counties <- us_counties %>%
  filter(STATEFP == "24")

trump_hogan_map_data <- md_counties %>%
  left_join(senate_pres_data, by = c("NAME" = "county"))

ggplot(trump_hogan_map_data) +
  geom_sf(aes(fill = trump_hogan_diff), color = "white") + 
  scale_fill_gradient(
    low = "lavender",
    high = "purple",
    na.value = "gray"
  ) +
  theme_minimal() +
  labs(
    title = "Trump vs. Hogan performance by county in 2024",
    subtitle = "Percentage difference between Trump's and Hogan's vote shares",
    fill = "Trump - Hogan %"
  ) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic"),
    legend.position = "bottom"
  )

```

**Q3** Make another map showing the difference between Larry Hogan's county-level performance this year and from his governor race in 2018, using percentages to compare the two as you did in Q2. You can use the dataframe you initially loaded for Q1, and you'll need to load the 2018 governor results in the data folder and join them to it before proceeding. Are there any counties where Hogan did better this year? How would you describe the map showing the difference?

**Hogan’s vote share declined in nearly every county compared to 2018. His major declines appeared to be in suburban and rural areas where he previously had stronger crossover support, which is pretty interesting and might be worth looking into further.**

```{r}
gov_18 <- read_csv("data/md_gov_county_18.csv") %>%
  clean_names()

gov_18 <- gov_18 %>%
  mutate(
    total_votes = hogan + jealous,
    hogan_pct_18 = hogan / total_votes * 100
  ) %>%
  select(county, hogan_pct_18)

hogan_compare <- senate_data %>%
  select(county, hogan_pct) %>%
  left_join(gov_18, by = "county") %>%
  mutate(hogan_diff = hogan_pct - hogan_pct_18)

hogan_map_data <- md_counties %>%
  left_join(hogan_compare, by = c("NAME" = "county"))

ggplot(hogan_map_data) +
  geom_sf(aes(fill = hogan_diff), color = "white") +
  scale_fill_gradient2(
    low = "purple", mid = "white", high = "green", midpoint = 0,
    na.value = "gray90"
  ) +
  theme_minimal() +
  labs(
    title = "Change in Hogan's county-level vote share (2018 to 2024)",
    subtitle = "Percentage point difference in Hogan's vote share between 2018 and 2024",
    fill = "2024 - 2018 %"
  ) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic"),
    legend.position = "bottom"
  )

```

**Q4.** Choose your own map adventure! In the data folder I've included Maryland county-level results for the abortion rights amendment vote this year, plus the 2020 presidential results by county and 2022 governor's results by county. Using one of those and at least one other contest for comparison, make a county-level map showing what you think is the most interesting aspect of that data, and describe the results in terms of a possible story. You can bring in other data, such as Census information, if you like.

**Support for the abortion rights amendment in 2024 was higher than Wes Moore’s 2022 vote count across several Maryland counties. The map highlights how abortion rights attracted broader support than the Democratic candidate did, including in areas that often vote Republican or swing between parties. This shows that more people who voted for other candidates for governor still support abortion rights regardless of stereotyped political party affiliation.**

```{r}
gov_22 <- read_csv("data/md_gov_county_22.csv") %>%
  clean_names()

gov_22 <- gov_22 %>%
  mutate(
    total_votes = moore + cox,
    moore_pct = moore / total_votes * 100
  ) %>%
  select(county, moore_pct)

abortion_compare <- abortion_24 %>%
  left_join(gov_22, by = "county") %>%
  mutate(diff_pct = for_pct - moore_pct)

abortion_map_data <- md_counties %>%
  left_join(abortion_compare, by = c("NAME" = "county"))

ggplot(abortion_map_data) +
  geom_sf(aes(fill = diff_pct), color = "white") +
  scale_fill_gradient(
    low = "lavender",
    high = "purple",
    na.value = "gray90"
  ) +
  labs(
    title = "Abortion Rights Support vs Moore's 2022 Vote Share",
    subtitle = "Difference between 2024 'For' % and Moore 2022 %",
    fill = "For - Moore %"
  )

```

-30-
